An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-09-22 paper.tex
Add some to-do notes
Add §5.2 on Influence, from ICWSM 2022 paper

\title{An Empirical Analysis of the Internet Engineering Task Force 
       with Computational Methods}
\author{Colin Perkins\\University of Glasgow}
\maketitle
%==================================================================================================
\begin{abstract}
  % Four sentences:
  %  - State the problem
  %  - Say why it's an interesting problem
  %  - Say what your solution achieves
  %  - Say what follows from your solution

@@

%==================================================================================================
\section{Trends in RFC Publication}

@@

@@

%==================================================================================================
\section{Organisational Dynamics of the IETF}

%--------------------------------------------------------------------------------------------------
\subsection{Characterising Participant Interactions}

@@

%..................................................................................................
\pb{Discussion of drafts:}
The mailing lists are, other than plenary or interim meetings, the primary
means for the discussion of drafts.  To measure how often drafts are
discussed, we next identify mentions of drafts in mailing list messages.
We therefore extract any mention of a draft (beginning \texttt{draft-}) or
RFC (i.e., ``RFC'' followed by a number).  Figure~\ref{fig:draft_mentions}
presents the number of drafts mentions in the emails per year. Separate
mentions of the same draft are counted as different mentions, as we want to
observe the entire volume of mentions.  We observe a strong increase in the
number of mentions over time.  This is largely driven by the growing number
of drafts being published. In fact, we find a Pearson correlation of 0.89
between the number of drafts published and the number of mentions. This
speaks to the influence of emails that mention drafts in driving draft
production.

@@

%--------------------------------------------------------------------------------------------------
\subsection{Influence}

@@

%--------------------------------------------------------------------------------------------------
\subsection{Hierarchy}
%..................................................................................................
% ICWSM 2024 paper
% - RQ1: Are higher levels of the hierarchy growing or becoming more
%   centralised? Are they associated with an increased domination of the
%   conversation? 
% - RQ2: What is the association between the organisation hierarchy and
%   general communication patterns?
% - RQ3: How do people with differing roles communicate with each other?
%   Does information flow “up” or “down” the hierarchy? 
% - RQ4: What is the impact that individuals have on their direct contacts’
%   activity over time? How does this vary across hierarchy level?

@@

>>>>

\title{An Empirical Analysis of the Internet Engineering Task Force 
       with Computational Methods}
\author{Colin Perkins \and
        Prashant Khare \and
        Mladen Karan \and
        Stephen McQuistin \and
        Waleed Iqbal \and
        Junaid Qadir \and
        Matthew Purver \and
        Patrick Healey \and
        Ignacio Castro \and
        Gareth Tyson}

\maketitle
%==================================================================================================
\begin{abstract}
  % Four sentences:
  %  - State the problem
  %  - Say why it's an interesting problem
  %  - Say what your solution achieves
  %  - Say what follows from your solution

@@

%==================================================================================================
\section{Trends in RFC Publication}
\label{sec:trends-documents}

@@

\todo{Add draft publication rate as mentioned in Section \ref{sec:org-dynamics:participants}}

@@

%==================================================================================================
\section{Organisational Dynamics of the IETF}
\label{sec:org-dynamics}

%--------------------------------------------------------------------------------------------------
\subsection{Interactions}
\label{sec:org-dynamics:participants}

@@

%..................................................................................................
\pb{Discussion of drafts:}
The mailing lists are, other than plenary or interim meetings, the primary
means for the discussion of drafts.  To measure how often drafts are
discussed, we next identify mentions of drafts in mailing list messages.
We therefore extract any mention of a draft (beginning \texttt{draft-}) or
RFC (i.e., ``RFC'' followed by a number).  Figure~\ref{fig:draft_mentions}
presents the number of drafts mentions in the emails per year. Separate
mentions of the same draft are counted as different mentions, as we want to
observe the entire volume of mentions.  We observe a strong increase in the
number of mentions over time.  This is largely driven by the growing number
of drafts being published. In fact, we find a Pearson correlation of 0.89
between the number of drafts published and the number of mentions. This
speaks to the influence of emails that mention drafts in driving draft
production.
\todo{Add discussion of number of drafts published over time in Section
\ref{sec:trends-documents}. Does this tie into the discussion of increasing
complexity of draft publication?}

@@

%--------------------------------------------------------------------------------------------------
\subsection{Influence}
\label{sec:org-dynamics:influence}

@@

% This is modified from the Introduction to the ICWSM 2022 paper:

As we showed in Section \ref{sec:org-dynamics:participants}, protocol
standardisation is an inherently social process with most day-to-day work
happening on public mailing lists, aided by meetings, video conference
calls, and open document and code repositories.  We are specifically
interested in better understanding how influence is distributed across
stakeholders and how it might affect the standardisation process.  This is
of critical societal importance: the IETF has a major impact on global
Internet technologies, and understanding the social processes involved
would give us insight into not only the driving forces behind
standardisation, but also its resilience to the loss of major participants.
In the following we consider the following \emph{research questions}: 
\begin{enumerate}
    \item How centralised is the active IETF community, and to what extent
      is it reliant on a small core of participants? 
    \item How do the most influential participants behave? 
    \item How does influence (determined by mailing list participation)
      relate to wider impacts throughout the IETF?
    \item Does the organisational affiliation of participants also
      influence the innovation (adoption of new work) within IETF?
\end{enumerate}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% ICWSM 2022 paper Section 3
\subsubsection{The IETF as a Social Graph}

\label{subsec:measuring_influence}

Much of the interaction between IETF participants occurs on public mailing
lists.  For each year, we build a social graph based on the \emph{active
community} of participants (nodes) who have interactions (edges) with any
other participant in the previous 5 years.

In Section \ref{sec:org-dynamics:participants}, we demonstrated that there
are three categories of participants in IETF - (a) \textit{young
contributors} -- who leave within one year of their first year mailing list
contribution; (b) \textit{mid-age contributors} -- participants who stay
active for up-to 5 years; and (c) \textit{senior contributors} -- who
remain active for more than 5 years. Following this, a 5 year period window
is chosen to observe interactions. There are no direct
participant-to-participant email exchanges in our dataset: the emails and
responses we capture are sent to the public mailing lists, and we observe
an interaction between two participants when one replies to an email sent
by another on any mailing list. This yields a social graph based on
1,049,793 emails (out of 2.1M) from 22,138 unique participants across 840
mailing lists.

%..................................................................................................
% ICWSM 2022 paper Section 3.1
\pb{Participation}
To examine the variation in participation between IETF participants, in
Figure~\ref{fig:degree_cdf} we plot the cumulative degree distribution of
the social graph -- i.e., the cumulative number of people a participant
interacts with.  We observe a core group that always interacts with
substantially more people than the rest of the community, with around 80\%
of the participants having degree  $<$15, but 5\%-10\% having degree $>$40.
However,  this difference decreases over time: in 2005-2009 around 6\% of
participants had degree $>$40, but by the period 2015-2019 this increases
to $\sim10\%$.  This shows that while there is always a core of more active
participants, the degree of participation has spread  out over time.
\todo{Repetitive with Section \ref{sec:org-dynamics:participants}?}

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/degree_CDF.pdf}
\caption{Cumulative degree distribution of the email graph for different year-periods.}
\label{fig:degree_cdf}
\end{figure}

To understand the structure of the community, and its reliance on specific
groups, we analyse the connected components of this graph. Each connected
component reflects a maximal set of nodes such that each pair of nodes is
connected by a path. We compute the size of the Largest Connected Component
(LCC) and the Number of Connected Components (NCC) for each year in
Figure~\ref{fig:lcc_size_yearly}.  The NCC peaks in 2003 before declining,
and broadly aligns with the variation in the number of meeting
attendees,\footnote{\url{https://datatracker.ietf.org/stats/meeting/overview/}}
suggesting a lag between participating for the first time and integrating
into the wider community.  In contrast, the size of the LCC increases until
2006 before stabilising.  Overall, we observe that the IETF community has
become less fragmented.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
\caption{Size of Largest Connected Component (LCC) and Number of Connected Components (NCC)}
\label{fig:lcc_size_yearly}
\end{figure}

%..................................................................................................
% ICWSM 2022 paper Section 3.1
\pb{Influence}
Relying on a small group to interconnect the community could undermine the
resilience of the IETF.  To study the influence of participants and their
role in interconnecting the wider community, we compute the betweenness
centrality of each participant in a given time
period~\cite{kourtellis2013identifying, weitzel2012measuring,
sole2014centrality}. We also considered other graph based influence metrics
such as eigenvector centrality: eigenvector centrality reflects the
importance of a node as per its neighbours, while betweenness centrality is
based on shortest paths which is independent of the influence of
neighbours.

However, we found a very strong correlation between the two measures, 
in line with similar experiments \cite{valente2008correlated,he2016correlation}:
the Spearman's rank correlation of participants ranked by betweenness
centrality and eigenvector centrality ranged between 0.51-0.72, with a
strong statistical significance ($p < 0.01$) for the period 2000-2019. We
therefore use just betweenness centrality in our analysis; this is often
studied and acknowledged as a measure of influence in social and complex
networks, particularly built on online communication
\cite{hagen2018crisis,chen2012identifying, ghalmane2019centrality}.

Figure~\ref{fig:purge_nodes_largest_connected_component_number_components}
then shows the effect of removing the most influential participants (in 1\%
increments from most to least, moving left-to-right on the x-axis) on the
size of the LCC.  Worryingly, we find that removal of just 20-25\% of the
most influential participants causes the LCC to shrink by 90\%.  However,
we also find that this impact has decreased over time: for instance, in
2000-2004, removing the top $\sim$5\% influential participants, reduces the
size of LCC by more than half, whereas in 2015-2019, it takes the removal
of the top $\sim$15\% of participants to have the same effect.  This shows
that the community has become \emph{more} cohesive and resilient over time,
and the IETF can now sustain a larger amount of churn while maintaining a
well-connected social graph.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lcc_proportion_granular_yearly.pdf}
\caption{Impact of removing participants by their influence on the size of the LCC.}
\label{fig:purge_nodes_largest_connected_component_number_components}
\end{figure}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% ICWSM 2022 paper Section 3.2
\subsubsection{Behaviour of Influential Participants}
\label{subsec:influencer_behaviour}

\todo{This is somewhat repetitive with Section \ref{sec:org-dynamics:participants}}

We now characterise the behaviour of the most influential participants, in
terms of the volume of emails sent, length of time active within the
community, and topics discussed.

%..................................................................................................
\pb{Email volume}
Figure~\ref{fig:emailcount_top_percentile} shows that each participant in
the top 10\% most influential participants sends on average around
0.05\%-0.08\% of total emails in a given period.  Collectively, the top
10\% most influential participants account for 43.75\%, on average, of the
total emails, a substantially larger proportion than the others.

This dynamic seems stable over time, making differences even more acute.
At the same time, the overall number of emails sent increases up to 2010
and remains roughly stable from then onward.  Along with the results from
\S\ref{subsec:measuring_influence}, this shows a worrying, if slowly
improving, dependence of the IETF on a small number of influential
participants.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
\caption{Proportion of emails sent (\%) by participants in each period 
(with 95\% confidence interval) according to their betweenness-centrality percentile (x-axis).
Y2-axis (purple) shows yearly count of emails.}
\label{fig:emailcount_top_percentile}
\end{figure}

%..................................................................................................
\pb{Cross-area review} 
In addition to sending more emails, we test if influential participants
engage with different parts of the community more. The work of the IETF is
divided into several areas (e.g., Applications \& Real-time, Security, and
Routing). This eases administration, but might also act as a barrier to
broader discussion and review. For example, WebRTC standards developed in
the Applications and Real-time area might contain elements that could
benefit from expertise from the Security area, but participants in one area
might not review work in another.

Figure~\ref{fig:areas_participation_top_percentile} shows the mean number
of areas where IETF participants are active (derived from the mailing lists
they use).  We see that influential participants engage in more areas of
the IETF, on average, and that cross-area engagement has improved over
time, indicating that the community itself has matured. This indicates that
influential participants benefit the IETF in enabling cross-area discussion
and review, and can bridge administrative divisions.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_avg_areas_top_percentile.pdf}
\caption{Mean number of areas participated in (with 95\% confidence
  interval) according to their betweenness-centrality percentile in the
  x-axis, ranked from  top (0-10) to bottom percentile (91-100).}
\label{fig:areas_participation_top_percentile}
\end{figure}

%..................................................................................................
\pb{Participation duration}
We next consider for how long participants remain associated with the IETF
(measured as the delta between first and last emails sent).
Figure~\ref{fig:age_top_percentile} shows the mean participation duration
distribution for participants, ranked by influence.  Participation duration
increases over time, with the most influential participants typically being
those who have been active for longer. While this might be expected, it
might be a good sign for the IETF community in that it implies that the
most influential participants are also the most experienced.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
\caption{Mean participation duration (with 95\% confidence interval)
  according to their betweenness-centrality percentile in the x-axis,
  ranked from  top (0-10) to bottom percentile (91-100).}
\label{fig:age_top_percentile}
\end{figure}

We extend this observation by considering what happens to participants once
they become influential. Figure \ref{fig:heatmap_top10_overlap} shows what
proportion of the top 10\% influential participants in any given year
(\emph{y-axis}) was also among the top 10\% influential participants in any
other year (\emph{x-axis}). This shows that a significant majority of
participants that become influential, continue to be influential for a
number of years: the top 10\% are influential for at least 6-7 years on
average.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
\caption{Yearly overlap (\%) of top 10\% influencers.}
\label{fig:heatmap_top10_overlap}
\end{figure}

Further, Figure \ref{fig:rate_new_influential_participants} shows that
breaking into the top 10\% of influencers requires an increasing number of
years of participation. This may be beneficial, showing that the IETF is
maturing and is capable of retaining influential and experienced
participants, but it may also point to an increasingly ossified structure
that is not welcoming to newcomers.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/new_influencers_top10_age_longevityInTop10.pdf}
\caption{Percentage of new influential participants breaking into top 10\%
  influencers (black); the average years of participation before entering
  top 10\% (purple); and minimum \#  years new influencers remain in top
  10\% (with 95\% confidence intervals).}
\label{fig:rate_new_influential_participants}
\end{figure}

%..................................................................................................
\pb{Topics of Discussion} 
The topics discussed on WG mailing lists are a good indicator of the focus
of technical work, and it might be expected that influential participants
will set the direction of that work.  To explore this, we use the Latent
Dirichlet Allocation (LDA) \cite{blei2003latent,hoffman2010online} model
from \texttt{gensim} \cite{rehurek_lrec} to induce 100 topics on the entire
set of email texts. 
Each topic is a distribution over words: e.g., a \emph{security} topic might
have high probability for \emph{cypher}, \emph{rsa}, \emph{auth}, and
related words. We can also use the model to obtain a vector for any input
text as a sparse distribution over all topics, e.g., finding that a text is
comprised of 30\% security and 70\% video streaming topics. With this, we
generate a vector for each participant in a given time period by
concatenating all messages sent by that participant in the period and
feeding it into the LDA model. The result can be loosely interpreted as a
distribution of topics on which each participant works.

Manual inspection of the induced topics reveals some interesting trends. We
find that topics related to \emph{routing} and \emph{email} protocols are
seeing a steady decline in popularity, while topics like \emph{streaming}
and \emph{cryptography} are becoming more prominent. This reflects wider
trends in standardisation, as these efforts have adjusted to the public's
increased awareness of privacy, especially in light of the Edward Snowden
leaks \cite{RFC7258}. We also find that as videoconferencing became
increasingly popular, so did the WebRTC protocol that frequently underpins
them.

We next measure the topical diversity of a participant by observing the
entropy of their topic vector, their \emph{topic entropy}, defined as
follows:
\begin{equation}
    \eta = - \sum_{i=1}^{|T|} p_i \log p_i \nonumber
\end{equation}
where $T = [p_1, ..., p_N]$ is a topic vector, which defines a probability
distribution over $N$ topics, each $p_i$ is a probability of the $i$-th
topic and $\sum_i p_i = 1$.

Figure \ref{fig:topic_diversity} shows topic entropy distributions of
participants in different time periods and influence percentiles. While we
initially experimented with measuring diversity by simply counting the
number of topics that account for the majority ($ \geq 95\% $) of a
participant's topic distribution probability mass, we found that most
participants engage in a relatively small number of topics, regardless of
influence. However, after measuring diversity using entropy, we identify
that the activity of the more influential participants tends to be more
evenly spread across the topics they participate in\footnote{An example
with 3 topics: a distribution of $[0.8, 0.1, 0.1]$ is less evenly spread
than $[0.3, 0.4, 0.3]$.}. This difference becomes more pronounced over
time, implying that increased topic entropy (i.e.,~more evenly participating
in different topics) is an increasingly salient property of influencers.
This is aligned with our earlier findings on cross-area review, showing the
growing number of areas in which influencers participate.

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/topic-div.png}
\caption{Mean topic entropy (with 95\% confidence interval) according to
  their betweenness-centrality percentile in the x-axis, ranked from  top
  (0-10) to bottom percentile (91-100).}
\label{fig:topic_diversity}
\end{figure}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Impact of Influential Participants}
\label{subsec:impact_influence}

% ICWSM 2022 paper Section 3.3

While these top participants are influential within the \emph{mailing
lists}, it is unclear how this influence translates into document
authorship and into gaining  leadership roles within  the organisation.

%..................................................................................................
\pb{Draft authorship}
The output of the IETF is technical standards and other documents published
in the RFC series. RFCs are developed by WGs from a sequence of Internet-Drafts,
with individuals acting as named authors.  Having identified the most
influential participants in the mailing list community, we can determine
whether or not these same participants are also the most active authors.

Figure~\ref{fig:drafts_initiation_top_percentile} shows the distribution of
the proportion of documents authored by mailing list participants, sorted
by their influence rank. We observe a growing proportion of the community
involved in draft authoring, skewed towards influential participants who
tend to write more drafts.  During 2010-2019, each participant in the top
10 percentile, authored 0.125\% to 0.175\% of the total number of drafts in
that period. We find that 32\%-42\% of the total drafts, in different years
during this period, were authored by the top 10\% most influential mailing
list participants.

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/documents_proportion_published_top_percentile.pdf}
\caption{Average drafts authored (with 95\% confidence interval) according
  to their betweenness-centrality percentile in the x-axis, ranked from
  top (0-10) to bottom percentile (91-100).  Right-y axis (purple) shows
  yearly drafts per year.}
\label{fig:drafts_initiation_top_percentile}
\end{figure}

%..................................................................................................
\pb{Co-authorship and email graph correlation}
Figure~\ref{fig:drafts_initiation_top_percentile} showed that influential
mailing list participants tend to author more drafts than others.
We next ask if they are also influencers in the draft co-authorship graph.

Thus, we create a draft co-authorship graph where each author is a node,
and draft co-authorship is an edge. We then measure influence with
betweenness centrality of the authors in each time period.
Table~\ref{tbl:correlation_sprmn} compares the top 20\% of influencers of
the mailing lists with those of the co-authorship graph.  We find a
significant overlap between both groups, ranging from 48.2\% to 67.2\%.
There is a significant ($p < 0.05$) positive correlation between the
rankings of the overlapping members of each community: participants that
are influential in the mailing lists are also likely to be influential in
draft authorship.

\begin{table}
\small 
\begin{tabular}{llllll}
\toprule
\multicolumn{6}{c}{Top 20\% draft authors \& all email participants}\\ \midrule

Years   & \multicolumn{2}{c}{Sub-Network Size} & Overlap  & \multicolumn{2}{c}{Spearman's $r_{s}$} \\ \cline{2-3}  \cline{5-6}
    &   Co-author & Email  &  &   $r_{s}$   & $p$-value \\ \midrule

2000-04   &   398 &   1390 &   48.24\%  &  .323 &   4.75e-06 \\
2005-09   &   427 &   1662 &   67.21\% &   .332   &   7.39e-09    \\
2010-14   &   728 &   1639 &   63.05\% &   .299   &   5.73e-11  \\
2015-19   &   915 &   1370 &   55.85\%  &   .337   &   4.13e-15  \\ %\hline

\bottomrule
\end{tabular}
\caption{Overlap in the co-author and email graph.}
\label{tbl:correlation_sprmn}
\end{table}

We also look at the top 20\% participants from the co-authorship network
who are \emph{not} part of the top 20\% influencers in the email network in
Figure \ref{fig:non_overlapping_coauthornetwork_authors_emailnetwork}.
We observe that not all the prolific authors are that engaged in the email
discussion: 40\%-50\% of non-overlapping authors are ranked between 20th to
40th percentile of influence in the email network. These non-overlapping
authors are typically more junior with respect to their participation
duration.

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
\caption{Percentage of Top 20\% authors that are not 20\% influencers
  according to their betweenness-centrality (email graph) percentile in the
  x-axis, ranked from  top (0-10) to bottom percentile (91-100).}
\label{fig:non_overlapping_coauthornetwork_authors_emailnetwork}
\end{figure}

%..................................................................................................
\pb{Leadership roles} 
WG chairs are selected from the community, and we might expect those
selected to be influential in the community.
Figure~\ref{fig:wg_chair_before_after} shows that this is indeed the case:
87.5\% of WG chairs are in the top 20\% of mailing list influencers and
67.4\% are in the top 20\% of document authors in the year before they
became chairs. This also shows the impact of taking up a leadership role:
influence in both the mailing list and authorship communities grows in the
year after participants become a WG chair for the first time.

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
 \caption{CDF of the percentile of betweenness-centrality of WG chairs in
  the email and co-authorship graph, one year before and after becoming
  chairs.}
\label{fig:wg_chair_before_after}
\end{figure}

%..................................................................................................
\pb{Organisations}  While participants in the IETF contribute as
individuals, they are usually affiliated with an organisation.  To study
the potential influence of organisations, for each draft we obtain the
authors' affiliations from the Datatracker. If this is not available, we
use the domain name of the authors' email addresses, which we map to the
relevant organisation (e.g., \texttt{@cisco.com} maps to Cisco). For generic
email addresses, such as \texttt{@gmail.com}, we use the participant name
if no affiliation is available.

Figure \ref{fig:coauthor_affiliation_network2019_otheryears} shows the ten
most frequent affiliated organisations of the top 20\% authors in the
period 2015-2019, and the number of authors affiliated with these
organisations in  each  period.  While the dominance of Cisco is clear,
other organisations, such as Huawei, have gained a larger presence over
time.  In general, a small number of organisations employ a significant
fraction of the influential participants in the standards process. If we
look at the period 2015-2019, for example, out of the 915 authors ranked in
the top 20\% of most influential authors, 342 belong to the ten most
frequently affiliated organisations, and nearly 253 are from Cisco, Huawei,
Ericsson, or Juniper alone.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
\caption{Authors in the most common affiliations of top 20\% authors
  (co-authorship graph) in 2015-2019.}
\label{fig:coauthor_affiliation_network2019_otheryears}
\end{figure}

While authors from the same organisation often co-author drafts together,
collaborations between authors from different organisations are also
common.  Figure~\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}
shows the most common collaborations between authors from the top 15 most
frequently affiliated organisations.  Collaborations between authors at
competing organisations, such as between authors from Cisco, Huawei,
Ericsson, and Juniper, occur frequently. We also computed that over the
years, joint collaboration (authors from multiple affiliations) has
increased: 772 co-authored drafts in the period 2000-2004 vs. 3083
co-authored drafts in the period 2015-2019.

\begin{figure}
\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/heatmap_2019_top10_affiliation_collaboration.pdf}
\caption{Number of drafts co-authored by top 15 most frequently affiliated
  organisations. We only include the top 20\% authors (co-authorship graph)
  of 2015-2019.}
\label{fig:heatmap_2019_coauthordraft_most_frequent_org}
\end{figure}

Further inspection of the collaboration trends reveals that the volume of
collaborations between authors from Huawei and other organisations appears
to be unaffected by the recent addition of Huawei to the U.S. Entity List
\cite{bis:2019:entity} as the trends in Figure~\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}
also hold for 2020-2021.\footnote{\tiny\url{https://mailarchive.ietf.org/arch/msg/ietf-announce/0ywjgSS4LlO0DaWDoLJLRHxJdUk/}}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Discussion}

% ICWSM 2022 paper section 4.3

We conclude that the IETF is still reasonably centralised, but that this
has improved over the years. However, removing around 20\%-25\% most
influential participants still fragments the entire network
(\S\ref{subsec:measuring_influence}, Figure
\ref{fig:purge_nodes_largest_connected_component_number_components}). The
most influential people are the ones with highest degree of engagements
within the community who also tend to be more involved in draft authoring
activities (\S\ref{subsec:influencer_behaviour}, Figure
\ref{fig:drafts_initiation_top_percentile}). However, we also observe that
over the years more people (even the ones less influential in the email
network) are getting involved in draft authoring activities (Figure
\ref{fig:drafts_initiation_top_percentile}). This shows that these
activities, while centralised to a certain degree, are still open to
contributions from the wider IETF community.

We observe that the influential participants show a much higher level of
engagement with the community (\S\ref{subsec:influencer_behaviour}). A
considerable proportion of total emails is sent by the top 10\% of
influential participants (Figure \ref{fig:emailcount_top_percentile}).
Compared to the rest of participants, they are active in more areas (Figure
\ref{fig:areas_participation_top_percentile}), are active within the IETF
for a longer period of time (Figure \ref{fig:age_top_percentile}), and
participate in a more diverse set of topics (Figure
\ref{fig:topic_diversity}). Finally, their influence extends from the
mailing lists to other activities such as draft authorship.

A significant overlap and correlation is observed between
influential authors from co-authorship network and email networks
(\S\ref{subsec:impact_influence}, Table \ref{tbl:correlation_sprmn}). This
shows that a large set of authors exhibit an ability to co-author drafts as
well as engage with participants on the email networks, thereby translating
their influence from email networks to co-authorship and vice versa.

A considerable portion of influential participants ($\sim$30\%) are
affiliated to one of the more prominent organisations (e.g., \emph{Cisco}
or \emph{Ericsson}) (\S\ref{subsec:impact_influence}, Figure
\ref{fig:coauthor_affiliation_network2019_otheryears}). Participants from
different organisations do considerably collaborate (Figure
\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}). Moreover, the
level of such collaboration is found to have increased with time.

%--------------------------------------------------------------------------------------------------
\subsection{Hierarchy}

% ICWSM 2024 paper
% - RQ1: Are higher levels of the hierarchy growing or becoming more
%   centralised? Are they associated with an increased domination of the
%   conversation? 
% - RQ2: What is the association between the organisation hierarchy and
%   general communication patterns?
% - RQ3: How do people with differing roles communicate with each other?
%   Does information flow “up” or “down” the hierarchy? 
% - RQ4: What is the impact that individuals have on their direct contacts’
%   activity over time? How does this vary across hierarchy level?

@@

% ICWSM 2022 paper section 4.3
The statistical analysis shows that influence of draft authors in the email
networks does impact the possibility of a draft getting adopted
(\S\ref{subsec:results}, Table \ref{tbl:res} \& \ref{tbl:resstat}). This
might hint at the ability of participants who hold a domain expertise to be
able to engage better with the community. Several WG chairs are already in
the top percentile influential category in both the email and co-authorship
networks before taking up these leadership roles, which further elevates
after taking up such leadership roles (Figure
\ref{fig:wg_chair_before_after}).

Most importantly the statistical analysis shows that being affiliated with
a prominent organisation positively impacts the chances of a draft in
getting adopted by a WG, thereby directly driving the innovation process
(Table \ref{tbl:res} \& \ref{tbl:resstat}).
